Finiteness in Cubical Type Theory

### 2020-06-07 figures/finite-classification.tex
diagrams etc

\begin{figure*}
  \centering
  \begin{tikzcd}[cramped, row sep=small]
    {} &
    {} \ar[ddddd, dash, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Non Discrete} &
    \ar[ddddd, dash, dashed, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Discrete} &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    &
    &
    \text{Manifest Enumerable}
    \ar[
      from=rr,
      bend left=15,
      crossing over,
      start anchor=south west,
      end anchor=south east
      ]
    \ar[
      rr,
      bend left=15,
      "\text{Discrete}" description,
      start anchor=north east,
      end anchor=north west
      ]
    \ar[ddd, crossing over]
    &
    &
    \text{Split Enumerable}
      \ar[d, xshift=-1ex, bend right=30]
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Ordered} &
    &
    &
    &
    \text{Manifest Bishop}
      \ar[u, xshift=1ex, bend right=30]
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, dashed, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Unordered} &
    {} &
    \text{Kuratowski}
    \ar[rr, crossing over, bend left=15, "\text{Discrete}" description, start anchor=north east, end anchor=north west]
    \ar[from=rr, crossing over, bend left=15, start anchor=south west, end anchor=south east]
    &
    {}
    &
    \text{Cardinal}
      \ar[from=uu, crossing over]
      \ar[uuu, crossing over, "\text{Ord}" description, end anchor=east, start
      anchor=east, in=-15, out=20]
    &
    \\
    & & & & &
    \\
    & \ar[rrrr, "\text{More Restrictive}" description, end anchor={[xshift=-6ex]}]
    &
    &
    &
    &
    \ar[uuuuuu, "\text{More Informative}" {description, near start}]
    {}
  \end{tikzcd}
  \caption{Classification of finiteness predicates according to whether they are
    discrete (imply decidable equality) and whether they imply a total order.}
  \label{finite-classification}
\end{figure*}%

>>>>

\begin{figure*}
  \centering
  \begin{tikzcd}[cramped, row sep=small, column sep=2.3em]
    {} &
    {} \ar[ddddd, dash, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Non Discrete} &
    \ar[ddddd, dash, dashed, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Discrete} &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Ordered} &
    &
    \text{\parbox{1.8cm}{\centering Manifest Enumerable}}
    \ar[
      from=rr,
      bend left=15,
      crossing over,
      start anchor=south west,
      end anchor={[yshift=-2ex]east}
      ]
    \ar[
      rr,
      bend left=15,
      "\text{Discrete}" description,
      start anchor={[yshift=2ex]east},
      end anchor=north west
      ]
    \ar[ddd, crossing over]
    &
    &
    \text{Split Enumerable}
      \ar[d, xshift=-1ex, bend right=30]
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    &
    &
    &
    &
    \text{Manifest Bishop}
      \ar[u, xshift=1ex, bend right=30]
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, dashed, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Unordered} &
    {} &
    \text{Kuratowski}
    \ar[rr, crossing over, bend left=15, "\text{Discrete}" description, start anchor=north east, end anchor=north west]
    \ar[from=rr, crossing over, bend left=15, start anchor=south west, end anchor=south east]
    &
    {}
    &
    \text{Cardinal}
      \ar[from=uu, crossing over]
      \ar[uuu, crossing over, "\text{Ord}" description, end anchor=east, start
      anchor=east, in=-15, out=20]
    &
    \\
    & & & & &
    \\
    & \ar[rrrr, "\text{More Restrictive}" description, end anchor={[xshift=-6ex]}]
    &
    &
    &
    &
    \ar[uuuuuu, "\text{More Informative}" {description, near start}]
    {}
  \end{tikzcd}
  \caption{Classification of finiteness predicates according to whether they are
    discrete (imply decidable equality) and whether they imply a total order.}
  \label{finite-classification}
\end{figure*}%

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../paper"
%%% End:

### 2020-06-07 sections/countdown.tex
diagrams etc

Take the following problem as an example:
\begin{gather*}
  \boxed{1} \boxed{3} \boxed{7} \boxed{10} \boxed{25} \boxed{50} \\
  \boxed{765} \tag{Target}
\end{gather*}
It has the following answer:
\begin{equation}
  3 \times ((7 \times (50 - 10)) - 25)
\end{equation}
Importantly, we do not have to use every number given to get to the target.
For our problem, we will use the operators \(\times\), \(+\), and \(-\).

@@

In order to find and prove countdown solutions, we will prove that the above
transformation is \emph{finite}: this will allow us to search exhaustively
through the available transformations, proving the existence or absence of a
transformation which is a solution.

@@

Let's first look at the transformation part.
As shown in figure~\ref{countdown-transform}, we can break the transformation
into four parts.
First, we notice that we don't have to use all of the given numbers in the
solution candidate: keeping or discarding input numbers, then, is the first step
of the transformation (fig.~\ref{countdown-filter}).
Next, we are allowed to rearrange the order of the numbers in our output: the
next step of the transformation, then, applies a permutation to the selected
numbers (fig.~\ref{countdown-permutation}).
Thirdly, we can choose the operators we're going to use: note that for \(n\)
chosen numbers, we can choose \(n - 1\) operators
(fig.~\ref{countdown-operators}).
Finally, we have to apply the binary operators with some parenthesising order
(fig.~\ref{countdown-parens}).

There are finitely many of each of these transformations: finitely many filters
for \(n\) objects, for instance, or finitely many tuples of \(n\) operators.
We are going to construct a type for each of these transformations, prove it
finite, and use that type to search.
The final, completed transformation will then be a dependent product of each
constituent step.

% \section{Filtering and Operators}
% \section{Enumerating Binary Trees}
% \section{Filtering Out Invalid Expressions with the Subobject Classifier}
% \section{Filtering Out Duplicates with Quotients}

>>>>

Here's an example puzzle:
\begin{gather*}
  \boxed{1} \boxed{3} \boxed{7} \boxed{10} \boxed{25} \boxed{50} \\
  \boxed{765} \tag{Target}
\end{gather*}
We'll allow the use of \(\times\), \(+\), and \(-\).
The answer is at the bottom of this page\footnotemark.

\footnotetext{\rotatebox[origin=c]{180}{Answer: \(3 \times ((7 \times (50 - 10)) - 25)\)}}

@@

We are going to define this transformation/filter pair as a type.
We will then show that this type is itself finite, and from there we will be
able to enumerate all possible solutions, derive the cardinality of the
solutions, and so on.

@@

Let's first look at the transformation aspect of the solution.
We have broken this transformation into four steps in
figure~\ref{countdown-transform}, each one feeding into the next.

First, we select which numbers in the input we are going to use in our solution
candidate: this is the filtering step of the transformation
(fig.~\ref{countdown-filter}).
Next we decide the order of the selected numbers in the output: this is the
permutation step of the transformation (fig.~\ref{countdown-permutation}).
We then select \(n-1\) operators to use with our \(n\) chosen numbers
(fig.~\ref{countdown-operators}).
Finally, we have to apply the binary operators with some parenthesising order
(fig.~\ref{countdown-parens}).

Each of these transformations can be represented by a finite object: a filter
for \(n\) items, for instance, can be represented as an \(n\)-tuple of bools.
The full transformation itself will be a product of each of these intermediate
steps, which itself is finite by closure under \(\Sigma\).

In this section we will use split-enumerability as our finiteness predicate.
Most of what we examine would work equally well with the other predicates, but
split enumerability is the most powerful and the simplest to define.
Since we will work only with set types, there won't be any need to look at
non-set predicates like manifest enumerability.
\section{Filtering and Operators}
The first component of the transformation is relatively simple: a filter.
We can represent this transformation as an \(n\)-tuple of bools.
The filter in fig.~\ref{countdown-filter}, for instance, could be represented
with the following:
\begin{equation}
  \langle 0 , 1 , 1 , 1 , 1 , 1  \rangle
\end{equation}
The proof that this type is finite is straightforward: by induction on \(n\), we
see that any \(n\)-tuple of finite things is itself finite.

At this point we can look forward slightly to an upcoming transformation and
prove it finite: the selection of \(n\) operators is \emph{also} constructed
using an \(n\)-tuple.
Clearly there are finitely many operators available to us, so an \(n\)-tuple of
operators is finite for the same reason as the filter.
\section{Permutations}
\section{Enumerating Binary Trees}
% \section{Filtering Out Invalid Expressions with the Subobject Classifier}
% \section{Filtering Out Duplicates with Quotients}
